---
title: "H2O Autoencoders for High-Dimensional Representation Learning"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(h2o)
```

## Import and prepare data

The project uses a relative data path so the analysis can be reproduced on another machine. See `data/README.md` for the expected file layout.

```{r}
set.seed(1821)

data_path <- file.path("data", "wave_functions.csv")
if (!file.exists(data_path)) {
  stop("Missing data/wave_functions.csv. See data/README.md for the expected layout.")
}

df <- fread(
  data_path,
  header = FALSE,
  nrows = 800000,
  skip = 1700000
) %>%
  select(1:113) %>%
  rename(Target = V113) %>%
  as_tibble()

if (ncol(df) != 113) {
  stop("Expected 112 predictors plus one target column.")
}

df <- df %>%
  mutate(
    Target = factor(
      Target,
      levels = c(1, 0),
      labels = c("Anomaly", "Smooth")
    )
  ) %>%
  slice_sample(prop = 0.003)
```

## Held-out test split

The external test partition is created first and is not used for preprocessing, architecture selection or sparsity tuning.

```{r}
set.seed(1821)
df_split <- initial_split(df, prop = 0.75, strata = Target)
df_train_raw <- training(df_split)
df_test_raw <- testing(df_split)
```

## Internal representation-selection split

Architecture and sparsity are selected on a second split made **inside the training partition**. Preprocessing for this search is learned from the representation-training subset only.

```{r}
set.seed(1821)
ae_split <- initial_split(df_train_raw, prop = 0.80, strata = Target)
ae_train_raw <- training(ae_split)
ae_valid_raw <- testing(ae_split)

selection_recipe <- recipe(Target ~ ., data = ae_train_raw) %>%
  step_YeoJohnson(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

selection_prep <- prep(selection_recipe, training = ae_train_raw)
ae_train <- bake(selection_prep, new_data = NULL)
ae_valid <- bake(selection_prep, new_data = ae_valid_raw)
```

## Start H2O

```{r}
h2o_instance <- h2o.init(max_mem_size = "5g")
h2o_instance

ae_train_h2o <- as.h2o(ae_train, destination_frame = "ae_train")
ae_valid_h2o <- as.h2o(ae_valid, destination_frame = "ae_valid")
selection_predictors <- setdiff(names(ae_train_h2o), "Target")
stopifnot(length(selection_predictors) > 1)
```

## Validation-based architecture and sparsity search

The H2O grid receives an explicit validation frame. With `nfolds = 0`, `h2o.getGrid(..., sort_by = "mse")` ranks models using validation metrics rather than training metrics. The best architecture and sparsity coefficient are then carried into the final model.

```{r}
architecture_grid <- list(
  hidden = list(
    c(50, 50, 4),
    c(4),
    c(50, 25),
    c(112, 56, 112, 4),
    c(112, 35, 4),
    c(112, 50, 4),
    c(112, 20, 4),
    c(112, 5, 4),
    c(150, 200, 4),
    c(200, 200, 200, 200, 50),
    c(200, 200, 200, 200, 200, 50, 25, 10)
  ),
  sparsity_beta = c(0.01, 0.05, 0.1, 0.2)
)

set.seed(1821)
search_result <- h2o.grid(
  algorithm = "deeplearning",
  x = selection_predictors,
  training_frame = ae_train_h2o,
  validation_frame = ae_valid_h2o,
  grid_id = "autoencoder_selection_grid",
  autoencoder = TRUE,
  activation = "Tanh",
  hyper_params = architecture_grid,
  search_criteria = list(
    strategy = "RandomDiscrete",
    max_models = 20,
    seed = 1821
  ),
  sparse = TRUE,
  average_activation = -0.1,
  standardize = FALSE,
  ignore_const_cols = FALSE,
  reproducible = TRUE,
  seed = 1821
)

selection_grid <- h2o.getGrid(
  "autoencoder_selection_grid",
  sort_by = "mse",
  decreasing = FALSE
)
selection_grid

best_search_model <- h2o.getModel(selection_grid@model_ids[[1]])
selected_hidden <- best_search_model@parameters$hidden
selected_sparsity_beta <- best_search_model@parameters$sparsity_beta

selected_hidden
selected_sparsity_beta
```

## Refit preprocessing on the complete training partition

Once hyperparameters have been selected, preprocessing is re-estimated using all rows in the overall training partition. The held-out test set is transformed with the fitted object and remains untouched by fitting.

```{r}
final_recipe <- recipe(Target ~ ., data = df_train_raw) %>%
  step_YeoJohnson(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

final_prep <- prep(final_recipe, training = df_train_raw)
df_train <- bake(final_prep, new_data = NULL)
df_test <- bake(final_prep, new_data = df_test_raw)

train_h2o <- as.h2o(df_train, destination_frame = "train_h2o")
test_h2o <- as.h2o(df_test, destination_frame = "test_h2o")

predictor_columns <- setdiff(names(train_h2o), "Target")
stopifnot(length(predictor_columns) > 1)
```

## Train the final sparse autoencoder

The selected architecture is retrained on the complete training partition. The recipe has already normalized predictors, so H2O's internal standardization is disabled.

```{r}
autoencoder <- h2o.deeplearning(
  x = predictor_columns,
  training_frame = train_h2o,
  autoencoder = TRUE,
  standardize = FALSE,
  shuffle_training_data = FALSE,
  reproducible = TRUE,
  seed = 1821,
  hidden = selected_hidden,
  activation = "Tanh",
  sparse = TRUE,
  average_activation = -0.1,
  sparsity_beta = selected_sparsity_beta
)

print(autoencoder)
```

## Inspect learned deep features

The last hidden layer is used as the learned representation. Its dimension therefore follows the architecture selected on the internal validation subset.

```{r}
embedding_layer <- length(selected_hidden)

train_deep_features <- h2o.deepfeatures(
  autoencoder,
  train_h2o,
  layer = embedding_layer
) %>%
  as.data.frame()

if (ncol(train_deep_features) >= 2) {
  train_deep_features %>%
    mutate(Target = df_train$Target) %>%
    ggplot(aes(x = .data[[names(train_deep_features)[1]]],
               y = .data[[names(train_deep_features)[2]]],
               color = Target)) +
    geom_point(alpha = 0.7) +
    labs(
      title = "First two learned deep features",
      x = "Deep feature 1",
      y = "Deep feature 2"
    ) +
    theme_minimal()
}
```

## PCA comparison

```{r}
pca_data <- recipe(Target ~ ., data = df_train_raw) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = 4) %>%
  prep(training = df_train_raw) %>%
  bake(new_data = NULL)

pca_data %>%
  ggplot(aes(PC1, PC2, color = Target)) +
  geom_point(alpha = 0.7) +
  labs(title = "PCA representation") +
  theme_minimal()
```

## Reconstruction error

```{r}
score_history <- as.data.frame(h2o.scoreHistory(autoencoder))

plot(
  score_history$training_mse,
  type = "l",
  xlab = "Scoring step",
  ylab = "Training MSE",
  main = "Final autoencoder training error"
)
```

## Export learned representations

```{r}
train_embedding <- h2o.deepfeatures(
  autoencoder,
  train_h2o,
  layer = embedding_layer
) %>%
  as.data.frame() %>%
  bind_cols(Target = df_train$Target)

test_embedding <- h2o.deepfeatures(
  autoencoder,
  test_h2o,
  layer = embedding_layer
) %>%
  as.data.frame() %>%
  bind_cols(Target = df_test$Target)

head(train_embedding)
head(test_embedding)
```

## Shut down H2O

```{r}
h2o.shutdown(prompt = FALSE)
```

## Session information

```{r}
sessionInfo()
```
