---
title: "H2O Autoencoders for High-Dimensional Representation Learning"
output: html_document
---

```{r setup, include=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(tidymodels)
library(h2o)
```

## Import and prepare data

The project uses a relative data path so the analysis can be reproduced on another machine. See `data/README.md` for the expected file layout.

```{r}
set.seed(1821)

data_path <- file.path("data", "wave_functions.csv")
if (!file.exists(data_path)) {
  stop("Missing data/wave_functions.csv. See data/README.md for the expected layout.")
}

df <- fread(
  data_path,
  header = FALSE,
  nrows = 800000,
  skip = 1700000
) %>%
  select(1:113) %>%
  rename(Target = V113) %>%
  as_tibble()

if (ncol(df) != 113) {
  stop("Expected 112 predictors plus one target column.")
}

df <- df %>%
  mutate(
    Target = factor(
      Target,
      levels = c(1, 0),
      labels = c("Anomaly", "Smooth")
    )
  ) %>%
  slice_sample(prop = 0.003)
```

## Train/test split and preprocessing

Preprocessing is learned from the training set only and applied unchanged to the held-out test set.

```{r}
set.seed(1821)
df_split <- initial_split(df, prop = 0.75, strata = Target)

df_train <- training(df_split)
df_test <- testing(df_split)

preprocess_rec <- recipe(Target ~ ., data = df_train) %>%
  step_YeoJohnson(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

preprocess_prep <- prep(preprocess_rec, training = df_train)

df_train <- bake(preprocess_prep, new_data = NULL)
df_test <- bake(preprocess_prep, new_data = df_test)
```

## Start H2O

```{r}
h2o_instance <- h2o.init(max_mem_size = "5g")
h2o_instance

train_h2o <- as.h2o(df_train, destination_frame = "train_h2o")
test_h2o <- as.h2o(df_test, destination_frame = "test_h2o")

predictor_columns <- setdiff(names(train_h2o), "Target")
stopifnot(length(predictor_columns) > 1)
```

## Train a sparse autoencoder

The recipe has already normalized the predictors, so H2O's internal standardization is disabled.

```{r}
autoencoder <- h2o.deeplearning(
  x = predictor_columns,
  training_frame = train_h2o,
  autoencoder = TRUE,
  standardize = FALSE,
  shuffle_training_data = FALSE,
  seed = 1821,
  hidden = c(200, 200, 200, 200, 50),
  activation = "Tanh",
  sparse = TRUE,
  sparsity_beta = 0.01
)

print(autoencoder)
```

## Inspect learned deep features

```{r}
train_deep_features <- h2o.deepfeatures(
  autoencoder,
  train_h2o,
  layer = 1
) %>%
  as.data.frame()

train_deep_features %>%
  mutate(Target = df_train$Target) %>%
  ggplot(aes(x = .data[[names(train_deep_features)[1]]],
             y = .data[[names(train_deep_features)[2]]],
             color = Target)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "First two learned deep features",
    x = "Deep feature 1",
    y = "Deep feature 2"
  ) +
  theme_minimal()
```

## PCA comparison

```{r}
pca_data <- recipe(Target ~ ., data = df_train) %>%
  step_pca(all_numeric_predictors(), num_comp = 4) %>%
  prep(training = df_train) %>%
  bake(new_data = NULL)

pca_data %>%
  ggplot(aes(PC1, PC2, color = Target)) +
  geom_point(alpha = 0.7) +
  labs(title = "PCA representation") +
  theme_minimal()
```

## Architecture grid search

```{r}
architecture_grid <- list(
  hidden = list(
    c(50, 50, 4),
    c(4),
    c(50, 25),
    c(112, 56, 112, 4),
    c(112, 35, 4),
    c(112, 50, 4),
    c(112, 20, 4),
    c(112, 5, 4),
    c(150, 200, 4),
    c(200, 200, 200, 200, 50),
    c(200, 200, 200, 200, 200, 50, 25, 10)
  )
)

architecture_search <- h2o.grid(
  algorithm = "deeplearning",
  x = predictor_columns,
  training_frame = train_h2o,
  grid_id = "autoencoder_architecture_grid",
  autoencoder = TRUE,
  activation = "Tanh",
  hyper_params = architecture_grid,
  sparse = TRUE,
  ignore_const_cols = FALSE,
  seed = 1821
)

h2o.getGrid(
  "autoencoder_architecture_grid",
  sort_by = "mse",
  decreasing = FALSE
)
```

## Sparsity search

```{r}
sparsity_grid <- list(
  sparsity_beta = c(0.01, 0.05, 0.1, 0.2)
)

sparsity_search <- h2o.grid(
  algorithm = "deeplearning",
  x = predictor_columns,
  training_frame = train_h2o,
  grid_id = "autoencoder_sparsity_grid",
  autoencoder = TRUE,
  hidden = c(200, 200, 200, 200, 50),
  activation = "Tanh",
  hyper_params = sparsity_grid,
  sparse = TRUE,
  average_activation = -0.1,
  ignore_const_cols = FALSE,
  seed = 1821
)

h2o.getGrid(
  "autoencoder_sparsity_grid",
  sort_by = "mse",
  decreasing = FALSE
)
```

## Reconstruction error

```{r}
score_history <- as.data.frame(h2o.scoreHistory(autoencoder))

plot(
  score_history$training_mse,
  type = "l",
  xlab = "Scoring step",
  ylab = "Training MSE",
  main = "Autoencoder training error"
)
```

## Export learned representations

The fifth hidden layer is the final 50-unit bottleneck in the chosen architecture.

```{r}
train_embedding <- h2o.deepfeatures(
  autoencoder,
  train_h2o,
  layer = 5
) %>%
  as.data.frame() %>%
  bind_cols(Target = df_train$Target)

test_embedding <- h2o.deepfeatures(
  autoencoder,
  test_h2o,
  layer = 5
) %>%
  as.data.frame() %>%
  bind_cols(Target = df_test$Target)

head(train_embedding)
head(test_embedding)
```

## Shut down H2O

```{r}
h2o.shutdown(prompt = FALSE)
```

## Session information

```{r}
sessionInfo()
```
